An Improved Method for Coupling Hydrodynamics with Reaction Networks and Nuclear Statistical Equilibrium

### 2021-09-27 simplified_sdc_intro/paper.tex
more styling

\newcommand{\Shydro}{{{\bf H}}}
\newcommand{\Hb}{{\bf H}}
\newcommand{\Rb}{{\bf R}}
\newcommand{\Rq}{{\bf R}}
\newcommand{\Adv}[1]{{\left [\boldsymbol{\mathcal{A}} \left(#1\right)\right]}}
\newcommand{\Advt}[1]{{\left [\boldsymbol{\mathcal{\tilde{A}}} \left(#1\right)\right]}}
\newcommand{\Advss}[1]{{\left [{\mathcal{{A}}} \left(#1\right)\right]}}
\newcommand{\Advs}[1]{\boldsymbol{\mathcal{A}} \left(#1\right)}

@@

We solve the Euler equations for compressible, reacting flow.  For ease
of exposition we describe the one-dimensional case;
multidimensional extensions are a straightforward modification to
include the CTU hydrodynamics scheme.
The system appears as:
\begin{align}
\label{eq:conserved_system}
\ddt{\rho} + \ddxs{(\rho u)} &= 0 \\
\ddt{(\rho X_k)} + \ddxs{(\rho X_k u)} &= \rho \omegadot(X_k) \\
\ddt{(\rho \alpha_k)} + \ddxs{(\rho \alpha_k u)} &= \rho \omegadot(\alpha_k) \\
\ddt{(\rho u)} + \ddxs{(\rho u^2 + p)} &= \rho g \\
\ddt{(\rho E)} + \ddxs{(\rho E u + p u)} &= \rho ug + \rho \Sdot 
\end{align}
where $\rho$ is the mass density, $u$ is velocity, $E$ is specific
total energy, $p$ is the pressure, and we carry nuclear species mass
fractions, $X_k$, and auxiliary composition variables, $\alpha_l$.  The specific total
energy relates to the specific internal energy, $e$, as $E = e + u^2/2$,
and we also separately evolve $e$, as part of a dual
energy formulation (see \citealt{bryan:1995,wdmergerI}):
\begin{equation}
\ddt{(\rho e)} + \ddxs{(\rho e u)} + p \ddx{u} = \rho \Sdot
\end{equation}
 The mass fractions are constrained to sum to 1, $\sum_k X_k = 1$, but
 no such constraint exists on the auxiliary variables.  Gravity,
 represented by the gravitational acceleration $g$ appears as a source
 term.  Finally, $\omegadot(X_k)$ is the creation rate for species
 $k$, $\omegadot(\alpha_l)$ is the creation rate for auxiliary
 composition variable $l$, and $\dot{S}$ is the energy generation rate
 per unit mass.  Aside from the $p\partial u/\partial x$ (``pdV'')
 term in the internal energy equation, this system is in conservative
 form with source terms.

@@

For convenience, we'll use $\Uc$ to represent the conserved state
and split the sources up into pure hydrodynamic and reactive parts:
\begin{equation}
  \Uc = \left ( \begin{array}{c}
           \rho \\
           \rho X_k \\
           \rho \alpha_l \\
           \rho u \\
           \rho E \\
           \rho e \end{array}\right )
  \qquad
  \Hb(\Uc) = \left ( \begin{array}{c}
    0 \\
    0 \\
    0 \\
    \rho g \\
    \rho u g \\
    0 \end{array} \right )
  \qquad
  \Rb(\Uc) = \left ( \begin{array}{c}
     0 \\
     \rho \omegadot(X_k) \\
     \rho \omegadot(\alpha_l) \\
     0 \\
     \rho \dot{S} \\
     \rho \dot{S} 
  \end{array} \right )
\end{equation}
We will represent the total source as $\Sc(\Uc)$:
\begin{equation}
  \Sc(\Uc) = \Hb(\Uc) + \Rb(\Uc).
\end{equation}
We note that the internal energy ``pdV'' work is not treated as a source
term but is instead constructed with the hydrodynamical fluxes are computed in the CTU method.

@@

Our simplified-SDC approach is based on the method described in \citet{SDC-old} and
a similar implementation in the \maestro\ simulation code.
We again start by considering the update
in integral form:
\begin{equation}
\Uc^{n+1} = \Uc^n + \int_t^{t+\Delta t} \left [ \Advs{\Uc} + \Rb(\Uc) \right ] dt,
\end{equation}
where $\Advs{\Uc}$ includes all terms in system
\ref{eq:conserved_system} except for the reactive source terms,
$\Rb(\Ub)$ (i.e., it includes the flux divergences, ``pdV'' term for
internal energy, and the pure hydrodynamical sources).  In this approach, we 
approximate the advective term
piecewise constant in time, using the value at the midpoint in time,
to achieve second-order accuracy, giving us:
\begin{equation}
\label{eq:integral:simplesdc}
\Uc^{n+1} = \Uc^n + \int_t^{t+\Delta t} \left \{ \Adv{\Uc}^{n+1/2} + \Rb(\Uc) \right \} dt
\end{equation}
The \castro\ version of the simplified-SDC algorithm
differs from the previous versions due to the need to do some operations
on the conserved variable state and some on the primitive variable
state.

@@

    \item construct the advective update terms, $\Advt{\Uc}^{n+1/2,(k)}_{i}$,
      by first ignoring the hydrodynamics sources,
      \begin{align}
        \Advt{\Uc}^{n+1/2,(k)}_{i} =
          &- \frac{\Fb^\xv(\Uc^{n+1/2,(k)}_{i+1/2}) - \Fb^\xv(\Uc^{n+1/2,(k)}_{i-1/2})}{\Delta x}
      \end{align}
    Now the conservative hydrodynamics source terms are computed by first updating to the
    new state with advection and the old-time source term applied for the full $\Delta t$ as: \MarginPar{little more complicated than this---we don't include the pdV source here}
    \begin{equation}
      \Uc^{\star\star} = \Uc^n + \Delta t \Advt{\Uc}^{n+1/2,(k)} + \Delta t \Hb(\Uc^n)
    \end{equation}
    %% first updating the density to the new state with advection only as:
    %% \begin{equation}
    %%   \rho^{\star\star} = \rho^n + \Delta t \Advt{\rho}^{n+1/2,(k)}
    %% \end{equation}
    %% then constructing the momentum source term:
    %% \begin{equation}
    %%   {\bf S}_{\rho\Ub}^{n+1/2} = \frac{1}{2} (\rho^n \gb^n + \rho^{\star\star} \gb^{n+1,(k-1)})
    %% \end{equation}
    %% Then updating the momentum with only advection as:
    %% \begin{equation}
    %%   (\rho \Ub)^{\star\star} = (\rho \Ub)^n + \Delta t \Advt{\rho\Ub}^{n+1/2,(k)} + \Delta t {\bf S}_{\rho\Ub}^{n+1/2}
    %% \end{equation}
    %% and finally constructing the energy source:
    %% \begin{equation}
    %%   S_{\rho E}^{n+1/2} = \frac{1}{2} \left [ (\rho\Ub)^n \cdot \gb^n + (\rho\Ub)^{\star\star} \cdot \gb^{n+1,(k-1)}
    %%     \right ]
    %% \end{equation}
    We then evaluate the source terms with $\Uc^{\star\star}$ and
    correct the advective term so that we have a time-centered
    source: \MarginPar{are we actually doing this correctly for SDC?}
    The final advective update term is then\footnote{For a source like gravity, this update can be done first for $\rho$ and then define the new momentum source using $\rho^{\star\star}$ and likewise for energy}:
    \begin{equation}
      \Adv{\Uc}^{n+1/2,(k)}_{i} = \Advt{\Uc}^{n+1/2,(k)}_{i} +
      \frac{\Delta t}{2} \left (\Shydro(\Uc^n) + \Shydro(\Uc^{\star\star}) \right )
    \end{equation}

@@

    \MarginPar{if $\Shydro$ depended on the outcome of the burn, we might need to do things differently here}

@@

    Note that our advection terms are piecewise
    constant \MarginPar{need to think about this more} approximations,
    so when we integrate our system, their contributions will be
    linear in time.  Since $(\rho X_k)$ sees a linear advective term,
    $\rho$ does as well.  $(\rho u)$ will also have linear
    contributions in time from advection.

>>>>

\newcommand{\Shydro}{{{\bf H}}}
\newcommand{\Hb}{{\bf H}}
\newcommand{\Rb}{{\bf R}}
\newcommand{\Rq}{{\bf R}}
\newcommand{\Adv}[1]{{\left [\boldsymbol{\mathcal{A}} \left(#1\right)\right]}}
\newcommand{\Advt}[1]{{\left [\boldsymbol{\mathcal{\tilde{A}}} \left(#1\right)\right]}}
\newcommand{\Advss}[1]{{\left [{\mathcal{{A}}} \left(#1\right)\right]}}
\newcommand{\Advsst}[1]{{\left [{\mathcal{\tilde{A}}} \left(#1\right)\right]}}
\newcommand{\Advs}[1]{\boldsymbol{\mathcal{A}} \left(#1\right)}

@@

We solve the Euler equations for compressible, reacting flow.  For ease
of exposition we describe the one-dimensional case;
multidimensional extensions are a straightforward modification to
include the CTU hydrodynamics scheme.  Our conserved variables are
\begin{equation}
  \Uc = \left ( \begin{array}{c}
           \rho \\
           \rho X_k \\
           \rho \alpha_l \\
           \rho u \\
           \rho E \\
           \rho e \end{array}\right )
\end{equation}
where $\rho$ is the mass density, $u$ is velocity, $E$ is specific
total energy, $p$ is the pressure, and we carry nuclear species mass
fractions, $X_k$, and auxiliary composition variables, $\alpha_l$.  The specific total
energy relates to the specific internal energy, $e$, as $E = e + u^2/2$,
and we also separately evolve $e$, as part of a dual
energy formulation (see \citealt{bryan:1995,wdmergerI}).
 The mass fractions are constrained to sum to 1, $\sum_k X_k = 1$, but
 no such constraint exists on the auxiliary variables. 
Defining the hydrodynamical fluxes:
\begin{equation}
  \Fb(\Uc) = \left ( \begin{array}{c}
         \rho u \\
         \rho X_k u \\
         \rho \alpha_l u \\
         \rho u^2 p \\
         (\rho E + p) u \\
         \rho e u \end{array}\right )
\end{equation}
We can write the system in conservative form for all state variables aside from $(\rho e)$ as:
\begin{equation}
  \ddt{\Uc} + \ddx{\Fb(\Uc)} = \Sc(\Uc)
\end{equation}
where for the special case of $(\rho e)$, we have an additional ``pdV'' term:
\begin{equation}
\ddt{(\rho e)} + \ddx{F(\rho e)} + p \ddx{u} = S(\rho e)
\end{equation}

We'll split the source term, $\Sc(\Uc)$ into pure hydrodynamic and reactive parts:
\begin{equation}
  \Hb(\Uc) = \left ( \begin{array}{c}
    0 \\
    0 \\
    0 \\
    \rho g \\
    \rho u g \\
    0 \end{array} \right )
  \qquad
  \Rb(\Uc) = \left ( \begin{array}{c}
     0 \\
     \rho \omegadot(X_k) \\
     \rho \omegadot(\alpha_l) \\
     0 \\
     \rho \dot{S} \\
     \rho \dot{S} 
  \end{array} \right )
\end{equation}
with 
\begin{equation}
  \Sc(\Uc) = \Hb(\Uc) + \Rb(\Uc).
\end{equation}
Here, gravity, represented by the gravitational acceleration $g$
appears as a pure hydrodynamical source term.  From reactions,
$\omegadot(X_k)$ is the creation rate for species $k$,
$\omegadot(\alpha_l)$ is the creation rate for auxiliary composition
variable $l$, and $\dot{S}$ is the energy generation rate per unit
mass.  We note that the internal energy ``pdV'' work is not treated as
a source term but is instead constructed with the hydrodynamical
fluxes are computed in the CTU method.

Aside from the $p\partial u/\partial x$ (``pdV'') term in the internal
energy equation, this system is in conservative form with source
terms.

We'll define the advective terms with hydrodynamic sources, $\Advs{\Uc}$ as
\begin{equation}
\Advs{\Uc} = -\ddx{\Fb(\Uc)} + \Hb(\Uc)
\end{equation}
for the general case, with the $(\rho e)$ component again having the extra
``pdV'' term:
\begin{equation}
\mathcal{A}(\rho e) = -\ddx{F(\rho e)} -p \ddx{u} + H(\rho e)
\end{equation}

@@

@@

Our simplified-SDC approach is based on the method described in \citet{SDC-old} and
a similar implementation in the \maestro\ simulation code.
We again start by considering the update
in integral form:
\begin{equation}
\Uc^{n+1} = \Uc^n + \int_t^{t+\Delta t} \left [ \Advs{\Uc} + \Rb(\Uc) \right ] dt,
\end{equation}
In this approach, we 
approximate the advective term
piecewise constant in time, using the value at the midpoint in time,
to achieve second-order accuracy, giving us:
\begin{equation}
\label{eq:integral:simplesdc}
\Uc^{n+1} = \Uc^n + \int_t^{t+\Delta t} \left \{ \Adv{\Uc}^{n+1/2} + \Rb(\Uc) \right \} dt
\end{equation}
The \castro\ version of the simplified-SDC algorithm
differs from the previous versions due to the need to do some operations
on the conserved variable state and some on the primitive variable
state.

@@

    \item construct the advective update terms, $\Advt{\Uc}^{n+1/2,(k)}_{i}$,
      by first ignoring the hydrodynamics sources,
      \begin{equation}
        \Advt{\Uc}^{n+1/2,(k)}_{i} =
          - \frac{\Fb^\xv(\Uc^{n+1/2,(k)}_{i+1/2}) - \Fb^\xv(\Uc^{n+1/2,(k)}_{i-1/2})}{\Delta x}
      \end{equation}
      for the general case, and
      \begin{align}
        \Advsst{\rho e}^{n+1/2,(k)}_{i} =
          &- \frac{F\left((\rho e)^{n+1/2,(k)}_{i+1/2}\right) - F\left((\rho e)^{n+1/2,(k)}_{i-1/2}\right)}{\Delta x} \\
          &- \left (\frac{p_{i+1/2}^{n+1/2,(k)} + p_{i-1/2}^{n+1/2,(k)}}{2} \right )\frac{u_{i+1/2}^{n+1/2,(k)} - u_{i-1/2}^{n+1/2,(k)}}{\Delta x}
      \end{align}
      for $(\rho e)$.

    Now the conservative hydrodynamics source terms are computed by first updating to the
    new state with advection and the old-time source term applied for the full $\Delta t$ as: 
    \begin{equation}
      \Uc^{\star\star} = \Uc^n + \Delta t \Advt{\Uc}^{n+1/2,(k)} + \Delta t \Hb(\Uc^n)
    \end{equation}
    %% first updating the density to the new state with advection only as:
    %% \begin{equation}
    %%   \rho^{\star\star} = \rho^n + \Delta t \Advt{\rho}^{n+1/2,(k)}
    %% \end{equation}
    %% then constructing the momentum source term:
    %% \begin{equation}
    %%   {\bf S}_{\rho\Ub}^{n+1/2} = \frac{1}{2} (\rho^n \gb^n + \rho^{\star\star} \gb^{n+1,(k-1)})
    %% \end{equation}
    %% Then updating the momentum with only advection as:
    %% \begin{equation}
    %%   (\rho \Ub)^{\star\star} = (\rho \Ub)^n + \Delta t \Advt{\rho\Ub}^{n+1/2,(k)} + \Delta t {\bf S}_{\rho\Ub}^{n+1/2}
    %% \end{equation}
    %% and finally constructing the energy source:
    %% \begin{equation}
    %%   S_{\rho E}^{n+1/2} = \frac{1}{2} \left [ (\rho\Ub)^n \cdot \gb^n + (\rho\Ub)^{\star\star} \cdot \gb^{n+1,(k-1)}
    %%     \right ]
    %% \end{equation}
    We then evaluate the source terms with $\Uc^{\star\star}$ and
    correct the advective term so that we have a time-centered
    source.
    The final advective update term is then\footnote{For a source like gravity, this update can be done first for $\rho$ and then define the new momentum source using $\rho^{\star\star}$ and likewise for energy}:
    \begin{equation}
      \Adv{\Uc}^{n+1/2,(k)}_{i} = \Advt{\Uc}^{n+1/2,(k)}_{i} +
      \frac{\Delta t}{2} \left (\Shydro(\Uc^n) + \Shydro(\Uc^{\star\star}) \right )
    \end{equation}

@@

    We note that this procedure works because the source terms here do
    not depend on the outcome of the reactions.

@@
